---
title: "Ecological Drivers of Stickleback Morphological Variation"
author: "John Berini"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  chunk_output_type: console
---

### Bring in our data and load the packages
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)   # Automatically loads dplyr, ggplot2, forcats, and lubridate
library(tidytext)

library(MixRF)
library(randomForest)
library(caret)
library(ranger)      
library(car)         
library(vegan)       
library(lmerTest)    
library(Hmisc)       
library(ggh4x)

# GitHub Safety Guard: Ensure local directories exist to prevent saving errors
if(!dir.exists("figures")) dir.create("figures")

# 1. Load Primary Datasets
sticks <- read.csv("data/stickle_morph_2.csv", header=TRUE) 
names(sticks)[names(sticks) == "Population"] <- "lake_code"
sticks$lake_code <- trimws(tolower(sticks$lake_code))

covs <- read.csv("data/rf_dat.csv", header=TRUE) 
covs$lake_code <- trimws(tolower(covs$lake_code))

sonde <- read.csv("data/VancouverSondeData2023.csv", header=TRUE)
zoop_raw <- read.csv("data/ZoopIDDataUpdated.csv", header=TRUE)
lake_mapping <- read.csv("data/lake_id_mapping.csv", header=TRUE)

# 2. Build the lookup dictionary
clean_lookup <- lake_mapping %>%
  mutate(
    join_key = trimws(tolower(original_value)),
    LakeID   = trimws(tolower(LakeID))
  ) %>%
  select(join_key, LakeID) %>%
  distinct()

# Map 'sticks' and 'covs' through the lookup dictionary
sticks <- sticks %>%
  mutate(join_key = lake_code) %>%
  left_join(clean_lookup, by = "join_key") %>%
  mutate(lake_code = coalesce(LakeID, lake_code)) %>%
  select(-join_key, -LakeID)

covs <- covs %>%
  mutate(join_key = lake_code) %>%
  left_join(clean_lookup, by = "join_key") %>%
  mutate(lake_code = coalesce(LakeID, lake_code)) %>%
  select(-join_key, -LakeID)

#add clean covariate labels

clean_labels <- c(
  "max depth est"     = "Max Depth (Est.) [m]",
  "max_depth_est"     = "Max Depth (Est.) [m]",
  "perimeter"         = "Lake Perimeter [m]",
  "mean basal area"   = "Forest Basal Area [m²/ha]",
  "mean_basal_area"   = "Forest Basal Area [m²/ha]",
  "mean crwn clsre"   = "Crown Closure [%]",
  "mean_crwn_clsre"   = "Crown Closure [%]",
  "mean proj ht"      = "Canopy Height [m]",
  "mean_proj_ht"      = "Canopy Height [m]",
  "mean sa proj ht"   = "Stand Height [m]",
  "mean_sa_proj_ht"   = "Stand Height [m]",
  "mean spp1 pc"      = "Dominant Sp. Cover [%]",
  "mean_spp1_pc"      = "Dominant Sp. Cover [%]",
  "mean sp3 cvr"      = "Tertiary Sp. Cover [%]",
  "mean_sp3_cvr"      = "Tertiary Sp. Cover [%]",
  "mean vert comp"    = "Vertical Complexity",
  "mean_vert_comp"    = "Vertical Complexity",
  "mean harvest yr"   = "Forest Harvest Year",
  "mean_harvest_yr"    = "Forest Harvest Year",
  "DEM vi"            = "Elevation [m]",
  "mean DEM vi"       = "Elevation [m]",
  "mean_DEM_vi"       = "Elevation [m]",
  "mean d2 ancient"   = "Dist. to Ancient Forest [m]",
  "mean_d2_ancient"   = "Dist. to Ancient Forest [m]",
  "mean d2 lake"      = "Dist. to Nearest Lake [m]",
  "mean_d2_lake"      = "Dist. to Nearest Lake [m]",
  "mean d2 ogdc"      = "Dist. to Unpaved Road [m]",
  "mean_d2_ogdc"      = "Dist. to Unpaved Road [m]",
  "mean d2 paved"     = "Dist. to Paved Road [m]",
  "mean_d2_paved"     = "Dist. to Paved Road [m]",
  "mean rd dnse paved"= "Paved Road Density [km/km²]",
  "mean_rd_dnse_paved"= "Paved Road Density [km/km²]",
  
  # Climate — Deficit & Moisture Index
  "mean CMD05"        = "Hargreaves Deficit (May) [mm]",
  "mean_CMD05"        = "Hargreaves Deficit (May) [mm]",
  "mean CMD08"        = "Hargreaves Deficit (Aug) [mm]",
  "mean_CMD08"        = "Hargreaves Deficit (Aug) [mm]",
  "mean CMD sp"       = "Hargreaves Deficit (Spring) [mm]",
  "mean_CMD_sp"       = "Hargreaves Deficit (Spring) [mm]",
  "mean CMI07"        = "Moisture Index (Jul) [mm]",
  "mean_CMI07"        = "Moisture Index (Jul) [mm]",
  "mean CMI08"        = "Moisture Index (Aug) [mm]",
  "mean_CMI08"        = "Moisture Index (Aug) [mm]",
  "mean CMI09"        = "Moisture Index (Sep) [mm]",
  "mean_CMI09"        = "Moisture Index (Sep) [mm]",
  "mean CMI10"        = "Moisture Index (Oct) [mm]",
  "mean_CMI10"        = "Moisture Index (Oct) [mm]",
  "mean CMI12"        = "Moisture Index (Dec) [mm]",
  "mean_CMI12"        = "Moisture Index (Dec) [mm]",
  
  # Climate — Degree Days
  "mean DD 0 05"      = "Degree Days >0°C (May)",
  "mean_DD_0_05"      = "Degree Days >0°C (May)",
  "mean_DD0_05"       = "Degree Days >0°C (May)",
  "mean DD 0 11"      = "Degree Days >0°C (Nov)",
  "mean_DD_0_11"      = "Degree Days >0°C (Nov)",
  "mean_DD0_11"       = "Degree Days >0°C (Nov)",
  "mean DD 5 06"      = "Degree Days >5°C (Jun)",
  "mean_DD_5_06"      = "Degree Days >5°C (Jun)",
  "mean_DD5_06"       = "Degree Days >5°C (Jun)",
  "mean DD 18 02"     = "Degree Days <18°C (Feb)",
  "mean_DD_18_02"     = "Degree Days <18°C (Feb)",
  "mean_DD18_02"      = "Degree Days <18°C (Feb)",
  "mean DD 18 06"     = "Degree Days <18°C (Jun)",
  "mean_DD_18_06"     = "Degree Days <18°C (Jun)",
  "mean_DD18_06"      = "Degree Days <18°C (Jun)",
  "mean DD 18 09"     = "Degree Days <18°C (Sep)",
  "mean_DD_18_09"     = "Degree Days <18°C (Sep)",
  "mean_DD18_09"      = "Degree Days <18°C (Sep)",
  "mean DD 18 at"     = "Degree Days <18°C (Autumn)",
  "mean_DD_18_at"     = "Degree Days <18°C (Autumn)",
  "mean_DD18_at"      = "Degree Days <18°C (Autumn)",
  "mean DD 18 sp"     = "Degree Days <18°C (Spring)",
  "mean_DD_18_sp"     = "Degree Days <18°C (Spring)",
  "mean_DD18_sp"      = "Degree Days <18°C (Spring)",
  "mean DD5 02"       = "Degree Days >5°C (Feb)",
  "mean_DD5_02"       = "Degree Days >5°C (Feb)",
  "mean DD5 11"       = "Degree Days >5°C (Nov)",
  "mean_DD5_11"       = "Degree Days >5°C (Nov)",
  
  # Climate — Reference Evaporation & Precip/Humidity
  "mean Eref09"       = "Hargreaves Ref. Evap. (Sep) [mm]",
  "mean_Eref09"       = "Hargreaves Ref. Evap. (Sep) [mm]",
  "mean Eref10"       = "Hargreaves Ref. Evap. (Oct) [mm]",
  "mean_Eref10"       = "Hargreaves Ref. Evap. (Oct) [mm]",
  "mean Eref at"      = "Hargreaves Ref. Evap. (Autumn) [mm]",
  "mean_Eref_at"      = "Hargreaves Ref. Evap. (Autumn) [mm]",
  "mean NFFD sm"      = "Frost-Free Days (Summer)",
  "mean_NFFD_sm"      = "Frost-Free Days (Summer)",
  "mean PAS sp"       = "Snow Precip. (Spring) [mm]",
  "mean_PAS_sp"       = "Snow Precip. (Spring) [mm]",
  "mean PAS05"        = "Snow Precip. (May) [mm]",
  "mean_PAS05"        = "Snow Precip. (May) [mm]",
  "mean PPT05"        = "Precipitation (May) [mm]",
  "mean_PPT05"        = "Precipitation (May) [mm]",
  "mean PPT06"        = "Precipitation (Jun) [mm]",
  "mean_PPT06"        = "Precipitation (Jun) [mm]",
  "mean PPT09"        = "Precipitation (Sep) [mm]",
  "mean_PPT09"        = "Precipitation (Sep) [mm]",
  "mean RH"           = "Relative Humidity [%]",
  "mean_RH"           = "Relative Humidity [%]",
  "mean RH sp"        = "Rel. Humidity (Spring) [%]",
  "mean_RH_sp"        = "Rel. Humidity (Spring) [%]",
  "mean RH wt"        = "Rel. Humidity (Winter) [%]",
  "mean_RH_wt"        = "Rel. Humidity (Winter) [%]",
  "mean RH01"         = "Rel. Humidity (Jan) [%]",
  "mean_RH01"         = "Rel. Humidity (Jan) [%]",
  "mean RH02"         = "Rel. Humidity (Feb) [%]",
  "mean_RH02"         = "Rel. Humidity (Feb) [%]",
  "mean RH03"         = "Rel. Humidity (Mar) [%]",
  "mean_RH03"         = "Rel. Humidity (Mar) [%]",
  "mean RH10"         = "Rel. Humidity (Oct) [%]",
  "mean_RH10"         = "Rel. Humidity (Oct) [%]",
  "mean RH11"         = "Rel. Humidity (Nov) [%]",
  "mean_RH11"         = "Rel. Humidity (Nov) [%]",
  "mean RH12"         = "Rel. Humidity (Dec) [%]",
  "mean_RH12"         = "Rel. Humidity (Dec) [%]",
  "mean Tave06"       = "Mean Temp (Jun) [°C]",
  "mean_Tave06"       = "Mean Temp (Jun) [°C]",
  "mean Tmax02"       = "Max Temp (Feb) [°C]",
  "mean_Tmax02"       = "Max Temp (Feb) [°C]",
  "mean Tmax09"       = "Max Temp (Sep) [°C]",
  "mean_Tmax09"       = "Max Temp (Sep) [°C]",
  "mean Tmax10"       = "Max Temp (Oct) [°C]",
  "mean_Tmax10"       = "Max Temp (Oct) [°C]",
  "mean Tmax12"       = "Max Temp (Dec) [°C]",
  "mean_Tmax12"       = "Max Temp (Dec) [°C]",
  "mean Tmax sp"      = "Max Temp (Spring) [°C]",
  "mean_Tmax_sp"      = "Max Temp (Spring) [°C]",
  
  # Water Column / Sonde
  "Mean fDOM RFU"     = "fDOM [RFU]",
  "Mean_fDOM_RFU"     = "fDOM [RFU]",
  "Mean pH"           = "pH",
  "Mean_pH"           = "pH",
  "SD DO Percent"     = "Dissolved O2 SD [%]",
  "SD_DO_Percent"     = "Dissolved O2 SD [%]",
  "SD fDOM RFU"       = "fDOM SD [RFU]",
  "SD_fDOM_RFU"       = "fDOM SD [RFU]",
  "SD Temp C"         = "Water Temp SD [°C]",
  "SD_Temp_C"         = "Water Temp SD [°C]"
)
```

#bridge and join datasets

```{r}
sonde_summary <- sonde %>%
  mutate(join_key = trimws(tolower(Site))) %>%
  left_join(clean_lookup, by = "join_key") %>%
  group_by(LakeID, Date) %>%
  summarise(
    Mean_Temp_C      = mean(TemperatureCelcius, na.rm = TRUE),
    SD_Temp_C        = sd(TemperatureCelcius, na.rm = TRUE),
    Mean_DO_Percent  = mean(PercentDissolvedOxygen, na.rm = TRUE),
    SD_DO_Percent    = sd(PercentDissolvedOxygen, na.rm = TRUE),
    Mean_DO_mgL      = mean(DissolvedOxygenmgL, na.rm = TRUE),
    SD_DO_mgL        = sd(DissolvedOxygenmgL, na.rm = TRUE),
    Mean_SPC_uScm    = mean(SPCuSpercm, na.rm = TRUE),
    SD_SPC_uScm      = sd(SPCuSpercm, na.rm = TRUE),
    Mean_pH          = mean(pH, na.rm = TRUE),
    SD_pH            = sd(pH, na.rm = TRUE),
    Mean_Phyco_RFU   = mean(PhycocyaninRFU, na.rm = TRUE),
    SD_Phyco_RFU     = sd(PhycocyaninRFU, na.rm = TRUE),
    Mean_Chl_RFU     = mean(ChlorophyllRFU, na.rm = TRUE),
    SD_Chl_RFU       = sd(ChlorophyllRFU, na.rm = TRUE),
    Mean_fDOM_RFU    = mean(fDOMRFU, na.rm = TRUE),
    SD_fDOM_RFU      = sd(fDOMRFU, na.rm = TRUE),
    .groups = "drop"
  )

zoop_standardized <- zoop_raw %>%
  mutate(join_key = trimws(tolower(LakeName))) %>%
  left_join(clean_lookup, by = "join_key") %>%
  select(-join_key)

master_environmental_file <- sonde_summary %>%
  rename(lake_code = LakeID) %>%
  group_by(lake_code) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

zoop_taxa_cols <- intersect(
  c("Calanoids", "Cyclopoids", "Nauplii", "Daphnia", "Bosmina", "Chydorus", 
    "Diaphanosoma", "Holopedium", "Chaoborus", "WaterMites", "Brachionus.Keratella", 
    "ConochilusColonies", "Asplanchna", "Ciliate", "Amoeboid", "Rotifers..other.", "Polyphemidae"),
  colnames(zoop_standardized)
)

zoop_scores <- data.frame(lake_code = character(), Zoop_Community_PC1 = numeric(), Zoop_Community_PC2 = numeric(), stringsAsFactors = FALSE)

zoop_pca_data <- zoop_standardized %>%
  filter(!is.na(LakeID) & LakeID != "") %>% 
  mutate(across(all_of(zoop_taxa_cols), ~ as.numeric(as.character(.x)))) %>%
  group_by(LakeID) %>%
  summarise(across(all_of(zoop_taxa_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

if(nrow(zoop_pca_data) > 0) {
  zoop_pca_df <- as.data.frame(zoop_pca_data)
  rownames(zoop_pca_df) <- zoop_pca_df$LakeID
  zoop_pca_df$LakeID <- NULL
  
  zoop_pca_df[is.na(zoop_pca_df)] <- 0
  row_totals <- rowSums(zoop_pca_df)
  zoop_pca_df <- zoop_pca_df[row_totals > 0, ]
  
  if(nrow(zoop_pca_df) > 0) {
    zoop_pca <- vegan::rda(vegan::decostand(zoop_pca_df, method = "hellinger"))
    
    # Extract and format variance explained by the zooplankton axes themselves
    zoop_importance <- summary(zoop_pca)$cont$importance
    zoop_variance_df <- data.frame(
      Axis = c("Zoop_PC1", "Zoop_PC2"),
      Eigenvalue = zoop_importance["Eigenvalue", c(1, 2)],
      Proportion_Explained = zoop_importance["Proportion Explained", c(1, 2)],
      Cumulative_Explained = zoop_importance["Cumulative Proportion", c(1, 2)]
    )
    write.csv(zoop_variance_df, "zooplankton_pca_variance_summary.csv", row.names = FALSE)
    
    extracted_scores <- as.data.frame(vegan::scores(zoop_pca, choices = c(1, 2), display = "sites"))
    colnames(extracted_scores) <- c("Zoop_Community_PC1", "Zoop_Community_PC2")
    
    zoop_scores <- extracted_scores
    zoop_scores$lake_code <- rownames(extracted_scores)
    rownames(zoop_scores) <- NULL
  }
}
```

covariate cleaning and imputation log
```{r}
combined_covs <- covs %>%
  left_join(master_environmental_file, by = "lake_code") %>%
  left_join(zoop_scores, by = "lake_code")

if("bedrock" %in% colnames(combined_covs)) combined_covs$bedrock <- as.factor(combined_covs$bedrock)
if("watershed" %in% colnames(combined_covs)) combined_covs$watershed <- as.factor(combined_covs$watershed)

# Count total numeric cells across all columns to find the denominator
total_numeric_cells <- combined_covs %>% 
  select(where(is.numeric)) %>% 
  is.na() %>% 
  length()

# Count how many cells are missing (NAs)
total_imputed_values <- combined_covs %>% 
  select(where(is.numeric)) %>% 
  is.na() %>% 
  sum()

# Calculate "percent missingness"
pct_imputed <- (total_imputed_values / total_numeric_cells) * 100

# Print a summary statement
message(paste0("\n>>> [Reviewer Metric]: Imputed ", total_imputed_values, 
               " missing numeric cells out of ", total_numeric_cells, 
               " total cells (", round(pct_imputed, 2), "% of the numeric matrix).\n"))

# Save imputation log to directory
imputation_log <- data.frame(
  Metric = c("Total_Numeric_Cells", "Total_Imputed_Cells", "Percent_Imputed"),
  Value  = c(total_numeric_cells, total_imputed_values, round(pct_imputed, 4)),
  stringsAsFactors = FALSE
)
write.csv(imputation_log, "metrics_imputation_summary.csv", row.names = FALSE)

# 1. Base lake-level imputation
combined_covs_clean <- combined_covs %>%
  mutate(across(where(is.numeric), ~ replace_na(as.numeric(.x), mean(as.numeric(.x), na.rm = TRUE)))) %>%
  mutate(across(where(is.factor), ~ fct_na_value_to_level(.x, level = "Missing")))

# 2. Dynamically identify all morphometric trait columns present in 'sticks'
trait_cols <- colnames(sticks)[!colnames(sticks) %in% c("lake_code", "Fish_ID", "ID", "Sample_ID", "Date")]

# 3. Expand out clean covariates to individual fish observations and drop missing traits
sticks_rf_dat <- sticks %>%
  left_join(combined_covs_clean, by = "lake_code") %>%
  drop_na(all_of(trait_cols))

# 4. Build predictor matrix for Random Forest (Fixing 'sticks_morph' reference)
rf_dat <- sticks_rf_dat %>%
  select(-all_of(trait_cols), -lake_code, -any_of("lake_name"), -any_of(zoop_taxa_cols)) %>%
  select(where(~ is.numeric(.) | is.factor(.)))

# 5. Omit complex multi-level factor variables (> 3 levels) that bias RF
rf_dat <- rf_dat %>%
  select(where(~ !is.factor(.x) || length(levels(.x)) <= 3))

# 6. Drop zero-variance numeric predictors safely
rf_dat <- rf_dat %>%
  select(where(~ !is.numeric(.x) || (is.numeric(.x) && var(.x, na.rm = TRUE) > 0)))
```

Random forest variable importance loop
```{r}
summary_list <- list()
set.seed(42) 

# 1. Define predictors from rf_dat and targets from trait_cols
predictor_names <- colnames(rf_dat)

# 2. Build forest_df directly using sticks_rf_dat
forest_df <- sticks_rf_dat %>% 
  select(all_of(c(predictor_names, trait_cols)))

# 3. Fit Random Forests for each dynamic trait column
for (t_name in trait_cols) {
  
  # Ensure complete cases for the specific trait being modeled
  model_data <- forest_df %>% 
    select(all_of(c(predictor_names, t_name))) %>% 
    drop_na(all_of(t_name))
  
  rf_fit <- ranger::ranger(
    formula = as.formula(paste(t_name, "~ .")),
    data = model_data,
    importance = "permutation", 
    num.trees = 1000,
    mtry = ceiling(length(predictor_names) / 3)
  )
  
  raw_imp <- rf_fit$variable.importance
  raw_imp[raw_imp < 0] <- 0 
  
  rel_importance = if(sum(raw_imp) > 0) (raw_imp / sum(raw_imp)) * 100 else rep(0, length(raw_imp))
  
  summary_list[[t_name]] <- data.frame(
    trait = t_name,
    variable = names(raw_imp),
    mean_rel_importance = as.numeric(rel_importance),
    stringsAsFactors = FALSE
  )
}

summary_table <- do.call(rbind, summary_list)
```

Variable Selection & Linear Model Filtering

```{r}
n_vars <- ncol(rf_dat)
imp_threshold <- 100 / n_vars

top_vars_dynamic <- summary_table %>%
  group_by(trait) %>%
  filter(mean_rel_importance > imp_threshold) %>%
  slice_max(order_by = mean_rel_importance, n = 10, with_ties = FALSE) %>% 
  ungroup()

final_plot_data <- list()

if(nrow(top_vars_dynamic) > 0) {
  for(i in seq_len(nrow(top_vars_dynamic))) {
    t_name <- top_vars_dynamic$trait[i]
    v_name <- top_vars_dynamic$variable[i]
      
    if(!v_name %in% colnames(rf_dat)) next
      
    # 1. Safely extract vectors and drop NAs
    y_raw <- sticks_rf_dat[[t_name]]
    x_raw <- rf_dat[[v_name]]
    
    # Preserve factor type if categorical; otherwise ensure numeric
    if(!is.factor(x_raw)) {
      x_raw <- as.numeric(as.character(x_raw))
    }
    y_raw <- as.numeric(as.character(y_raw))

    loop_df <- data.frame(
      y_val = y_raw,
      x_val = x_raw,
      stringsAsFactors = FALSE
    ) %>% 
      drop_na(y_val, x_val)
    
    # Check data volume and variance before scaling
    if(nrow(loop_df) < 5) next
    if(!is.factor(loop_df$x_val) && var(loop_df$x_val, na.rm = TRUE) == 0) next
    
    # 2. Safely fit initial model for Cook's Distance check
    init_fit <- tryCatch({
      if(is.factor(loop_df$x_val)) {
        lm(y_val ~ x_val, data = loop_df)
      } else {
        lm(y_val ~ scale(x_val), data = loop_df)
      }
    }, error = function(e) NULL)
    
    if(is.null(init_fit)) next
    
    # 3. Apply Cook's Distance filter
    cd <- cooks.distance(init_fit)
    leverage_threshold <- 4 / nrow(loop_df)
    
    # Retain non-NA Cook's distance points below the threshold
    valid_cd_indices <- which(!is.na(cd) & cd <= leverage_threshold)
    loop_df <- loop_df[valid_cd_indices, ]
    
    # Re-verify data volume and variance post-filtering
    if(nrow(loop_df) < 5) next
    if(!is.factor(loop_df$x_val) && var(loop_df$x_val, na.rm = TRUE) == 0) next
    
    # 4. Final Model Fitting with tryCatch safety
    fit_res <- tryCatch({
      if(is.factor(loop_df$x_val)) {
        fit <- lm(y_val ~ x_val, data = loop_df)
        anova_summary <- anova(fit)
        stats <- as.data.frame(summary(fit)$coefficients)
        
        if(nrow(stats) < 2) return(NULL)
        
        list(
          p_val   = anova_summary["x_val", "Pr(>F)"],
          est_val = stats[2, "Estimate"],
          std_err = stats[2, "Std. Error"],
          t_val   = stats[2, "t value"]
        )
      } else {
        fit <- lm(y_val ~ scale(x_val), data = loop_df)
        stats <- as.data.frame(summary(fit)$coefficients)
        
        if(nrow(stats) < 2) return(NULL)
        
        list(
          p_val   = stats[2, "Pr(>|t|)"],
          est_val = stats[2, "Estimate"],
          std_err = stats[2, "Std. Error"],
          t_val   = stats[2, "t value"]
        )
      }
    }, error = function(e) NULL)
    
    if(is.null(fit_res)) next
    
    p_val   <- fit_res$p_val
    est_val <- fit_res$est_val
    std_err <- fit_res$std_err
    t_val   <- fit_res$t_val
      
    if(!is.na(p_val) && p_val < 0.05) { 
      stats_row <- data.frame(
        estimate  = est_val,
        std_error = std_err,
        t_value   = t_val,
        p_value   = p_val,
        variable  = v_name,
        trait     = t_name,
        is_factor = is.factor(loop_df$x_val),
        stringsAsFactors = FALSE
      )
      
      stats_row$sig_tier = case_when(
        stats_row$p_value < 0.001 ~ "***",
        stats_row$p_value < 0.01  ~ "**",
        stats_row$p_value < 0.05  ~ "*"
      )
      
      stats_row$category <- case_when(
        grepl("^Mean_|^SD_", v_name) ~ "Water Column (Sonde)",
        grepl("Zoop_Community_", v_name) ~ "Zooplankton Food Supply",
        TRUE                          ~ "Landscape & Climate"
      )
      
      final_plot_data[[length(final_plot_data) + 1]] <- stats_row
    }
  }
}
```

Visualize significant environmental drivers 
```{r visualize-drivers}
# Ensure plot_df_final exists and is populated
if (exists("final_plot_data") && length(final_plot_data) > 0) {
  plot_df_final <- do.call(rbind, final_plot_data)
}

# Build the plot 
if (exists("plot_df_final") && !is.null(plot_df_final) && nrow(plot_df_final) > 0) {

  p_coeff <- ggplot(
    plot_df_final, 
    aes(
      y = tidytext::reorder_within(variable, abs(estimate), trait),
      x = estimate,
      color = category
    )
  ) +
    geom_vline(xintercept = 0, linetype = "solid", color = "gray60", linewidth = 0.5) +
    geom_errorbar(
      data = subset(plot_df_final, variable != "(No significant covariates)"),
      aes(xmin = estimate - (1.96 * std_error),
          xmax = estimate + (1.96 * std_error)),
      width = 0.3, linewidth = 0.6
    ) +
    geom_point(
      data = subset(plot_df_final, variable != "(No significant covariates)"),
      size = 1.8
    ) +
    geom_text(
      data = subset(plot_df_final, variable != "(No significant covariates)"),
      aes(label = sig_tier,
          x = if_else(estimate >= 0, estimate + (1.96 * std_error), estimate - (1.96 * std_error)),
          hjust = if_else(estimate >= 0, -0.3, 1.3)),
      vjust = 0.4, size = 3, fontface = "bold", show.legend = FALSE
    ) +
    geom_text(
      data = subset(plot_df_final, variable == "(No significant covariates)"),
      aes(y = tidytext::reorder_within(variable, abs(estimate), trait),
          x = 0, label = "No significant relationships found"),
      color = "firebrick", fontface = "italic", size = 3, show.legend = FALSE
    ) +
    scale_y_discrete(labels = function(x) {
      clean_x <- gsub("___.*$", "", x)
      sapply(clean_x, function(val) {
        if (exists("clean_labels") && val %in% names(clean_labels)) {
          clean_labels[val]
        } else {
          stringr::str_squish(stringr::str_replace_all(val, "[._]", " "))
        }
      })
    }) +
    facet_wrap2(~ trait, scales = "free", ncol = 2, axes = "all", remove_labels = "none",
                labeller = labeller(trait = function(x) gsub("\\.", " ", x))) +
    scale_color_manual(
      values = c(
        "Landscape & Climate"     = "#2c3e50",
        "Water Column (Sonde)"    = "#2980b9",
        "Zooplankton Food Supply" = "#27ae60",
        "None"                    = "gray60"
      ),
      drop = FALSE
    ) +
    labs(
      title = "Significant Drivers of Stickleback Morphology",
      subtitle = "Standardized Fixed-Effects Slopes (Beta \u00b1 95% CI) | Linear Regression Architecture",
      y = "Model Predictors",
      x = "Standardized Effect Size (Continuous & Factor Indicators)",
      color = "Covariate Source",
      caption = "Significance levels: * p < 0.05, ** p < 0.01, *** p < 0.001"
    ) +
    theme_minimal(base_size = 9) +
    theme(
      panel.border     = element_rect(color = "gray85", fill = NA, linewidth = 0.5),
      strip.background = element_rect(fill = "gray96", color = "gray85"),
      strip.text       = element_text(face = "bold", size = 8.5),
      panel.spacing.x  = unit(1.0, "lines"),
      panel.spacing.y  = unit(0.6, "lines"),
      axis.text.y      = element_text(size = 6.5, color = "gray20"),
      axis.text.x      = element_text(size = 7.5),
      axis.title       = element_text(size = 9, face = "bold"),
      legend.position  = "bottom",
      legend.text      = element_text(size = 8)
    )

  # Display plot
  print(p_coeff)

  # Ensure figures directory exists before exporting as PNG
  if (!dir.exists("figures")) dir.create("figures", recursive = TRUE)

  ggsave(
    filename = "figures/Figure_1_Stickleback_Drivers.png",
    plot     = p_coeff,
    width    = 8.5,
    height   = 11,
    units    = "in",
    dpi      = 300
  )
}
```


Build summary table of results
```{r}
# Build summary table directly from validated model metrics
if (nrow(plot_df_final) > 0) {
  pub_table_data <- plot_df_final %>%
    filter(variable != "(No significant covariates)") %>%
    transmute(
      Trait     = trait,
      Category  = category,
      Predictor = variable,
      Beta      = estimate,
      SE        = std_error,
      t_stat    = t_value,
      p_val     = p_value
    )
} else {
  pub_table_data <- data.frame(
    Trait = character(), Category = character(), Predictor = character(),
    Beta = numeric(), SE = numeric(), t_stat = numeric(), p_val = numeric(), 
    stringsAsFactors = FALSE
  )
}

# Flag traits that lack statistically significant drivers (using trait_cols)
missing_traits <- setdiff(trait_cols, unique(pub_table_data$Trait))

if(length(missing_traits) > 0) {
  empty_rows <- data.frame(
    Trait     = missing_traits, 
    Category  = "None", 
    Predictor = "(No significant predictors found)",
    Beta      = NA_real_, 
    SE        = NA_real_, 
    t_stat    = NA_real_, 
    p_val     = NA_real_, 
    stringsAsFactors = FALSE
  )
  pub_table_data <- dplyr::bind_rows(pub_table_data, empty_rows)
}

# Helper function to match clean labels dynamically
format_clean_name <- function(x) {
  sapply(x, function(val) {
    if (is.na(val) || val == "(No significant predictors found)") return(val)
    if (val %in% names(clean_labels)) return(clean_labels[val])
    
    val_space <- chartr("_", " ", val)
    if (val_space %in% names(clean_labels)) return(clean_labels[val_space])
    
    val_underscore <- chartr(" ", "_", val)
    if (val_underscore %in% names(clean_labels)) return(clean_labels[val_underscore])
    
    # Fallback cleanup
    stringr::str_squish(stringr::str_replace_all(val, "[._]", " "))
  }, USE.NAMES = FALSE)
}

# Format columns for publication readiness
pub_table_formatted <- pub_table_data %>%
  mutate(
    Predictor = format_clean_name(Predictor),
    Beta      = if_else(is.na(Beta), "-", sprintf("%.3f", Beta)),
    SE        = if_else(is.na(SE), "-", sprintf("%.3f", SE)),
    t_stat    = if_else(is.na(t_stat), "-", sprintf("%.2f", t_stat)),
    p_val     = case_when(
      is.na(p_val)          ~ "-",
      as.numeric(p_val) < 0.001 ~ "< 0.001",
      TRUE                  ~ sprintf("%.3f", as.numeric(p_val))
    )
  )

# Output Table
knitr::kable(pub_table_formatted, caption = "Table 1. Linear regression models evaluating environmental drivers.")
write.csv(pub_table_formatted, "stickleback_regression_table.csv", row.names = FALSE)
```

### Robustness validation: Distance-Based Redundancy Analysis (dbRDA)
```{r, results='asis'}
# 1. Build matrix using dynamic trait_cols
morph_data_raw <- sticks_rf_dat %>% select(all_of(trait_cols))

valid_rows <- which(complete.cases(morph_data_raw) & complete.cases(rf_dat))

if(length(valid_rows) > 5) {
  morph_matrix_scaled <- scale(as.matrix(morph_data_raw[valid_rows, ]))
  rf_matrix_clean     <- rf_dat[valid_rows, ]
  
  morph_distance <- vegan::vegdist(morph_matrix_scaled, method = "euclidean")
  
  if("Zoop_Community_PC1" %in% colnames(rf_matrix_clean)) {
    dbrda_model <- vegan::dbrda(morph_distance ~ Zoop_Community_PC1 + Zoop_Community_PC2, data = rf_matrix_clean)
    
    r2_raw <- vegan::RsquareAdj(dbrda_model)
    global_permutation_test <- vegan::anova.cca(dbrda_model, permutations = 999)
    marginal_axis_test      <- vegan::anova.cca(dbrda_model, by = "margin", permutations = 999)
    
    # Extract global model P-value for Table 2a
    global_p_val <- global_permutation_test$`Pr(>F)`[1]
    formatted_p_val <- if(is.na(global_p_val)) {
      "-"
    } else if(global_p_val < 0.0001) {
      "< 0.0001"
    } else {
      sprintf("%.4f", global_p_val)
    }

    # Build Table 2a
    r2_df <- data.frame(
      Metric = c("R-Squared (Raw)", "Adjusted R-Squared", "Model Significance (p-value)"),
      Value  = c(sprintf("%.4f", r2_raw$r.squared), 
                 sprintf("%.4f", r2_raw$adj.r.squared), 
                 formatted_p_val),
      stringsAsFactors = FALSE
    )
    
    # Print and save Table 2a
    print(knitr::kable(r2_df, 
                       caption = "Table 2a. Morphological variance explained by the zooplankton community matrix (dbRDA R² and p-value).",
                       align = "lc"))
    write.csv(r2_df, "Table_2a_dbrda_variance_explained_table.csv", row.names = FALSE)
    cat("\n\n") 
    
    format_cca_table <- function(cca_res) {
      df_out <- as.data.frame(cca_res) %>% rownames_to_column(var = "Effect")
      
      # Standardize variance, statistic, and p-value column names
      var_col  <- intersect(c("Variance", "SumOfSqs", "Sum of Sqs"), colnames(df_out))[1]
      p_col    <- intersect(c("Pr(>F)", "Pr..F."), colnames(df_out))[1]
      f_col    <- intersect(c("F", "F.Mean"), colnames(df_out))[1]
      
      df_out <- df_out %>%
        mutate(
          Variance = if(!is.na(var_col)) sprintf("%.4f", .data[[var_col]]) else "-",
          F_stat   = if(!is.na(f_col)) ifelse(is.na(.data[[f_col]]), "-", sprintf("%.4f", .data[[f_col]])) else "-",
          p_val    = if(!is.na(p_col)) {
            sapply(.data[[p_col]], function(pv) {
              if (is.na(pv)) return("-")
              if (pv < 0.0001) return("< 0.0001")
              return(sprintf("%.4f", pv))
            })
          } else {
            "-"
          }
        ) %>%
        select(Effect, Df, Variance, F = F_stat, `Pr(>F)` = p_val)
      
      return(df_out)
    }
    
    # Format and save Global Model Table (Table 2b)
    global_formatted <- format_cca_table(global_permutation_test)
    print(knitr::kable(global_formatted, 
                       caption = "Table 2b. Global non-parametric permutation test for the overall dbRDA model.",
                       align = "lcccc"))
    write.csv(global_formatted, "Table_2b_dbrda_global_permutation_table.csv", row.names = FALSE)
    cat("\n\n")
    
    # Format and save Marginal Axis Table (Table 2c)
    marginal_formatted <- format_cca_table(marginal_axis_test)
    colnames(marginal_formatted)[1] <- "Axis"
    
    print(knitr::kable(marginal_formatted, 
                       caption = "Table 2c. Marginal permutation tests evaluating individual zooplankton PCA community axes.",
                       align = "lcccc"))
    write.csv(marginal_formatted, "Table_2c_dbrda_marginal_axis_table.csv", row.names = FALSE)
    
  } else {
    cat("\n WARNING: 'Zoop_Community_PC1' was not found in your predictor dataframe. Ensure the PCA step ran successfully.\n")
  }
} else {
  cat("\nInsufficient matching row indices available to run the multivariate validation matrix.\n")
}
```

Ecosystem gradient plot
```{r}
if (exists("final_plot_data") && length(final_plot_data) > 0) {
  
  # Rebuild plot_df_final from loop outputs if missing
  plot_df_final <- do.call(rbind, final_plot_data)
  
  # Ensure all traits are tracked
  all_traits <- trait_cols
  missing_traits <- setdiff(all_traits, unique(plot_df_final$trait))
  
  if (length(missing_traits) > 0) {
    empty_df <- data.frame(
      estimate = 0, std_error = 0, t_value = 0, p_value = 1,
      variable = "(No significant covariates)", trait = missing_traits,
      is_factor = FALSE, sig_tier = "", category = "None",
      stringsAsFactors = FALSE
    )
    plot_df_final <- bind_rows(plot_df_final, empty_df)
  }

  # Pull unique significant predictors AND their modeled categories
  sig_vars_info <- plot_df_final %>%
    filter(variable != "(No significant covariates)") %>%
    select(variable, category) %>%
    distinct()

  if (nrow(sig_vars_info) > 0) {
    
    # Extract matching raw numeric column names from rf_dat
    all_rf_cols <- colnames(rf_dat)
    
    # Safely align factor dummy labels back to rf_dat columns
    var_mapping <- do.call(rbind, lapply(seq_len(nrow(sig_vars_info)), function(i) {
      v <- sig_vars_info$variable[i]
      cat_val <- sig_vars_info$category[i]
      
      matched_col <- NA_character_
      if (v %in% all_rf_cols) {
        matched_col <- v
      } else {
        base_matches <- all_rf_cols[sapply(all_rf_cols, function(col) grepl(paste0("^", col), v))]
        if (length(base_matches) > 0) matched_col <- base_matches[1]
      }
      
      data.frame(raw_col = matched_col, category = cat_val, stringsAsFactors = FALSE)
    })) %>%
      drop_na(raw_col) %>%
      distinct()
    
    # Filter for numeric metrics only
    numeric_sig_vars <- rf_dat %>%
      select(all_of(intersect(var_mapping$raw_col, all_rf_cols))) %>%
      select(where(is.numeric)) %>%
      colnames()

    if (length(numeric_sig_vars) > 0) {
      
      # Pivot long and join exact modeled categories
      variation_long_df <- rf_dat %>%
        select(all_of(numeric_sig_vars)) %>%
        pivot_longer(cols = everything(), names_to = "raw_col", values_to = "raw_value") %>%
        drop_na(raw_value) %>%
        left_join(var_mapping, by = "raw_col") %>%
        rename(variable = raw_col)

      p_grad <- ggplot(variation_long_df, aes(x = "", y = raw_value, fill = category)) +
        geom_boxplot(alpha = 0.25, outlier.shape = NA, color = "gray30", width = 0.5, linewidth = 0.4) +
        geom_point(aes(color = category),
                   position = position_jitter(width = 0.2, height = 0, seed = 42),
                   alpha = 0.60, size = 1.2) +
        facet_wrap(~ variable, scales = "free_y", ncol = 5,
                   labeller = labeller(variable = function(x) {
                     sapply(x, function(val) {
                       if (exists("clean_labels") && val %in% names(clean_labels)) return(clean_labels[val])
                       stringr::str_squish(stringr::str_replace_all(val, "[._]", " "))
                     })
                   })) +
        scale_fill_manual(values = c(
          "Landscape & Climate"     = "#2c3e50",
          "Water Column (Sonde)"    = "#2980b9",
          "Zooplankton Food Supply" = "#27ae60"
        ), drop = FALSE) +
        scale_color_manual(values = c(
          "Landscape & Climate"     = "#2c3e50",
          "Water Column (Sonde)"    = "#2980b9",
          "Zooplankton Food Supply" = "#27ae60"
        ), drop = FALSE) +
        labs(
          title = "Ecosystem Gradient Profile: Variation Across Sample Sites",
          subtitle = "Distributions of significant phenotypic predictors; points mark individual observations",
          x = NULL,
          y = "Observed Environmental Metric Value (Raw Scales)",
          color = "Predictor Classification",
          fill = "Predictor Classification"
        ) +
        theme_minimal(base_size = 10) +
        theme(
          panel.border       = element_rect(color = "gray92", fill = NA, linewidth = 0.4),
          panel.grid.minor   = element_blank(),
          panel.grid.major.x = element_blank(),
          strip.background   = element_rect(fill = "gray97", color = "gray92"),
          strip.text         = element_text(face = "bold", size = 7, color = "gray20"),
          axis.text.y        = element_text(size = 6.5, color = "black"),
          axis.text.x        = element_blank(),
          axis.ticks.x       = element_blank(),
          panel.spacing.x    = unit(0.6, "lines"),
          panel.spacing.y    = unit(0.6, "lines"),
          legend.position    = "bottom",
          legend.margin      = ggplot2::margin(t = 2),
          legend.background  = element_rect(fill = "white", color = "gray95"),
          plot.title         = element_text(face = "bold", size = 12),
          plot.subtitle      = element_text(color = "gray40", size = 9, margin = ggplot2::margin(b = 8))
        ) +
        guides(color = guide_legend(nrow = 1), fill = guide_legend(nrow = 1))

      print(p_grad)

      if (!dir.exists("figures")) dir.create("figures", recursive = TRUE)

      ggsave("figures/Figure 2_Ecosystem_Gradients.png",
             plot = p_grad,
             width = 12,
             height = 9,
             units = "in",
             dpi = 300)
    }
  }
}
```

